Visualizing demographic evolution using geographically inconsistent census data

### 2019-10-10 paper/intro.tex
last paper review piccard 1.0 GEAN

The standard approach then is to \emph{geographically harmonise} the data. This
involves interpolating existing measurements into a common set of
regions~\citep{Logan2014,Hallisey2017,Allen2018}. \revision{1.18}{Recent
computational tools have somewhat simplified this
process~\citep{rey2018spatio}}, but it still involves non-trivial questions:
which geometry to use as target, how to apportion the variables, or how to
combine data from different sources. Further, these question do not necessarily
have optimal answers. Indeed, regardless of how well this process is performed,
it still introduces errors~\citep{Logan2016}, even when additional data is
provided~\citep{eicher2001dasymetric}. Essentially, harmonisation generates
\emph{artificial data points} that can potentially lead to inaccurate results,
even though they are seldom interpreted as such. Nevertheless, because there has
been no viable alternative, and the results often appear plausible, these
concerns are overlooked. The result is that the harmonisation approach is
virtually mandatory in the current literature: \emph{"(...) tract-by-tract
comparison is not possible unless data from 2000 is interpolated to 2010
boundaries (...)"}~\citep{Dmowska2017}, \emph{"(...) This limits cross-year
comparison since data are not representative of the same spatial units. (...)"}
~\citep{Allen2018}.

\revision{1.1}{In some cases, harmonisation might seem methodologically
adequate. If a study is focused on a small subregion that is fixed in time, limiting
and modifying the data to fit that specific region of interest sounds not only
reasonable, but recommended, even if it comes at the cost of producing artificial data points. However, this view is limited in that it ignores interactions beyond the fixed
borders: each different portion of the area of interest may belong to larger
homogeneous processes beyond these arbitrarily defined borders. This effect,
along with the creation of artificial points, is caused by the geometry mismatch
between the data and the region of interest. Instead of forcing one set of
borders into another, our method sidesteps this issue by changing how the data
is represented and interpreted. It therefore allows researchers to examine changes in both the composition and extent of local areas.

The main contribution of this paper is a method for longitudinal geographical
data representation and processing that works with the original data by
leveraging a network based representation. It enables direct comparison and
temporal regionalisation without the need to match arbitrarily defined borders.
Even in the case of a small, temporally fixed, region of interest, our approach
leads to a better understanding of its dynamics by exploring the related
data-driven homogeneous regions. }

@@

\revision{2.6.4}{While didactic, this example is not realistic because each
building is visible in Figure~\ref{fig:intuition}, so we know each individual of
the tabulation area, which is never the case. Further, real measurements are
seldom as homogeneous and noise-free as this artificial example.  In this
example, one possible harmonisation would split region A  in two to provide
different ancestors to C ($A_C$) and D ($A_D$). Then $A_D$ would not be
considered when analysing the evolution of C, nor $A_C$ for $D$. This is
inaccurate, because $A$ was the original state for both evolutions, and we
should not assume that these processes are disconnected.} Either way, by
splitting and merging the data to fit arbitrary borders, the harmonisation
process increases the distance between the data and reality by creating
artificial data points.

@@

Instead, we propose a network-based representation. A \emph{network} (also
called a \emph{graph}) is a collection of entities (nodes) that are related to
each other (edges). In this case, each different aggregation area is represented
as a node and we connect nodes that have overlapping geographical areas\revision{1.2}{ in
different times or are neighbours in the same time}, leading to the network
illustrated on the right of Figure~\ref{fig:intuition}. By partitioning the
network into connected nodes that are similar, we are effectively identifying
clusters in the spatio-temporal data, as illustrated by the colours of the nodes
on the right side of Figure~\ref{fig:intuition}. Further, all the possible paths
of change can be obtained by computing sequences of nodes over time, in this
case: (A, C, F), (A, D, F), (A, D, G), and (B, E, H). This representation is
also suited for geographically consistent regions, as illustrated by the stable
park in this example, and is therefore a generalisation of the traditional
paradigm.

>>>>

The standard approach then is to \emph{geographically harmonise} the data. This
involves interpolating existing measurements into a common set of
regions~\citep{Logan2014,Hallisey2017,Allen2018}. Recent
computational tools have somewhat simplified this
process~\citep{rey2018spatio}, but it still involves non-trivial questions:
which geometry to use as target, how to apportion the variables, or how to
combine data from different sources. Further, these question do not necessarily
have optimal answers. Indeed, regardless of how well this process is performed,
it still introduces errors~\citep{Logan2016}, even when additional data is
provided~\citep{eicher2001dasymetric}. Essentially, harmonisation generates
\emph{artificial data points} that can potentially lead to inaccurate results,
even though they are seldom interpreted as such. Nevertheless, because there has
been no viable alternative, and the results often appear plausible, these
concerns are overlooked. The result is that the harmonisation approach is
virtually mandatory in the current literature: \emph{"(...) tract-by-tract
comparison is not possible unless data from 2000 is interpolated to 2010
boundaries (...)"}~\citep{Dmowska2017}, \emph{"(...) This limits cross-year
comparison since data are not representative of the same spatial units. (...)"}
~\citep{Allen2018}.

In some cases, harmonisation might seem methodologically adequate. If a study is
focused on a small subregion that is fixed in time, limiting and modifying the
data to fit that specific region of interest sounds not only reasonable, but
recommended, even if it comes at the cost of producing artificial data points.
However, this view is limited in that it ignores interactions beyond the fixed
borders: each different portion of the area of interest may belong to larger
homogeneous processes beyond these arbitrarily defined borders. This effect,
along with the creation of artificial points, is caused by the geometry mismatch
between the data and the region of interest. Instead of forcing one set of
borders into another, our method sidesteps this issue by changing how the data
is represented and interpreted. It therefore allows researchers to examine
changes in both the composition and extent of local areas.

The main contribution of this paper is a method for longitudinal geographical
data representation and processing that works with the original data by
leveraging a network based representation. It enables direct comparison and
temporal regionalisation without the need to match arbitrarily defined borders.
Even in the case of a small, temporally fixed, region of interest, our approach
leads to a better understanding of its dynamics by exploring the related
data-driven homogeneous regions.

@@

While didactic, this example is not realistic because each building is visible
in Figure~\ref{fig:intuition}, so we know each individual of the tabulation
area, which is never the case. Further, real measurements are seldom as
homogeneous and noise-free as this artificial example.  In this example, one
possible harmonisation would split region A  in two to provide different
ancestors to C ($A_C$) and D ($A_D$). Then $A_D$ would not be considered when
analysing the evolution of C, nor $A_C$ for $D$. This is inaccurate, because $A$
was the original state for both evolutions, and we should not assume that these
processes are disconnected. Either way, by splitting and merging the data to
fit arbitrary borders, the harmonisation process increases the distance between
the data and reality by creating artificial data points.

@@

Instead, we propose a network-based representation. A \emph{network} (also
called a \emph{graph}) is a collection of entities (nodes) that are related to
each other (edges). In this case, each different aggregation area is represented
as a node and we connect nodes that have overlapping geographical areas in
different times or are neighbours in the same time, leading to the network
illustrated on the right of Figure~\ref{fig:intuition}. By partitioning the
network into connected nodes that are similar, we are effectively identifying
clusters in the spatio-temporal data, as illustrated by the colours of the nodes
on the right side of Figure~\ref{fig:intuition}. Further, all the possible paths
of change can be obtained by computing sequences of nodes over time, in this
case: (A, C, F), (A, D, F), (A, D, G), and (B, E, H). This representation is
also suited for geographically consistent regions, as illustrated by the stable
park in this example, and is therefore a generalisation of the traditional
paradigm.

### 2019-10-10 paper/method.tex
last paper review piccard 1.0 GEAN

\begin{figure}
    \centering 
    \includegraphics[width=0.975\linewidth]{overview.pdf}
    \caption{Overview of the proposed method. 
        \textbf{(a)} A network is generated from the original data,
        encoding the changing geographical information. 
        \textbf{(b)} The network is partitioned into an hierarchy. 
        \textbf{(c)} The characteristics and evolution of the clusters are then
        visually represented.
        \label{fig:overview}}
\end{figure}

@@

Each area of each census year is represented as a node, and edges are placed
between nodes if the corresponding regions share geographic borders in the same
year. Further, edges are placed between nodes if the corresponding regions
belong to sequential years and \revision{1.6/2.6.3}{their geometries intersect.
To avoid spurious connections caused by small fluctuations, one of the
geometries is slightly shrunk before the intersection, using a buffer of -1e6.
While the result would not change considerably by adding more connections
between nearby regions, these extra connections would greatly increase the
computational cost. More importantly, while weights will be associated with
these edges before they are processed, they are not derived from the geometry,
but from the data. The actual intersection area is not considered in this
representation. } This approach leads to a single network representing the whole
spatio-temporal space of the data.

\subsection{Geographic content clustering}
Having tied the regions together into a network, we can now partition it to
identify similar sets of regions. \revision{2.3}{In essence, we are performing
temporal regionalisation by applying a clustering method over this
spatio-temporal network.}

@@

Once the distances are associated to the edges, we use watershed
cuts~\citep{Cousty2009} to create an initial clustering, which is then refined
into a hierarchy using the Sorted Maximal Matching (SMM)~\citep{markus2017}. The
initial watershed step is performed to create an initial clustering and reduce
the running time of the SMM. We introduced one new parameter to this method: a
maximum distance threshold for the merges, to avoid the early merging of
outliers. We refer the reader to the original paper~\citep{markus2017} for more
details, including a complete performance evaluation using several metrics.
\revision{1.5/2.2}{We chose this algorithm because it is fast,
simple, and easily customisable, but our methodology should work with any
hierarchical network clustering algorithm. However, we caution against the use
of single-linkage methods due to their tendency to form chain-like clusters with
elements that are similar pairwise, but significantly different on the ends.
That is also the reason we did not use a Watershed hierarchy
directly~\citep{najman2011equivalence}.}

Each resulting cluster is contiguous in the network. This means that two
similar, but non-contiguous, sets of areas will be classified into two different
clusters, which can be counter-intuitive. To overcome this issue, we
\emph{augment} the network with two new edges per node from a nearest neighbours
graph~\citep{scikit-learn} using only the distances between the histograms.
These edges connect nodes with similar content, if they are not already
connected, providing a path for the algorithm to group similar nodes.
\revision{2.4}{The regions connected by those edges will be merged on the first
stages of the clustering, since the nodes they connect are, by definition, as
similar as possible\footnote{This assertion relies on the euclidean distance and
its relationship with the space where k-nn operates.}, leaving the remaining
steps of the hierarchy to be determined only by the geographical edges. We
experimented with different numbers of augmentation edges, but the results were
not consistent, since the distribution of the edges is data dependent. Adding
two edges per node was the smallest number of edges that led to stable and
consistent results in the scenarios available in our prototype. Since the
problem of balancing the data space with the geographical space is relevant for
geographical data analysis, this idea potentially warrants further exploration,
beyond the scope of this work.}

@@

We also use these statistical measurements to discover what characterises each
cluster, that is, what makes it different from the others.  We define the
\emph{relevance} of a variable of an aspect based on the distance between the
\revision{1.8}{interquartile} ranges (IQR) of the clusters in the same hierarchical level. If
the IQRs overlap for all clusters, that variable is not relevant to the
characterisation of the cluster, but if the IQRs are distant, it means that this
specific range of values is something that only occurs in this cluster.
Examining IQRs therefore provides users a straightforward visual method for
determining what variables most clearly define a given cluster.

@@

\subsection{Clusters and trajectories}
While the partition of the data into different clusters helps the user to
understand what groups exist and where they are, we are also interested in the
evolution of these groups. To examine this process of evolution directly, we
introduce the concepts\revision{2.6.4}{ of \emph{temporal paths} and \emph{trajectories}.

We call a temporal path any sequence of nodes in our representation network such
that the temporal information associated with the nodes only increases. For
instance, in Figure~\ref{fig:intuition}, the sequences ACF, ADF, ADG, and BEH
are temporal paths. With harmonised data, as illustrated by the path BEF, the
time-series of each region would form a temporal path, each node would be
connected only to its older and newer versions, and would belong to only one
temporal path. Since our data is not harmonised, more connections are allowed
and each node can belong to an arbitrary number of paths.

@@

Since each node in the sequence that forms the temporal paths has an associated
cluster, we can classify the paths based on the sequence of clusters. We call
each unique sequence of clusters present in this result a \emph{trajectory}.
Regions on the same trajectory had the same sequence of clusters, therefore had
similar temporal evolution.}

@@

\begin{figure*}
    \centering 
    \includegraphics[width=0.975\linewidth]{ui.pdf}
    \caption{Initial interface of our method showing the demographic evolution of Chicago and identifying the objectives of each plot.
        \textbf{(what)}: Cluster overview illustrating the most relevant aspect for each cluster. 
        \textbf{(where)}: Maps illustrating the geographical location, both as an overall summary and for each year.
        \textbf{(how)}: Tabular and visual summary of how the regions classification changed over time.
        \textbf{(who)}: Demographic details for the currently selected data.
        \label{fig:ui}}
\end{figure*}

@@

The cluster overview panel, marked as \emph{what} in Figure~\ref{fig:ui},
displays the most relevant aspects for each cluster, based on the distance
between the IQRs, as detailed in Section~\ref{sec:relevance}.\revision{2.5}{This
feature allows the user to quickly understand what exists in this region and
which aspect makes each cluster unique, without going into excessive detail. In
Chicago's example, we expect race to be relevant in general, and indeed, it is
recognised as the most relevant aspect for three of the four most distinct
clusters, with the fourth being education level. As we increase the number of
clusters, they get more nuanced, requiring more than one aspect for their proper
characterisation, as each cluster is divided into its composing sub-clusters.
The \emph{View all} button opens a new panel where all aspects are included,
while the chevron at the side lets the user expand each cluster separately,
allowing the exploration of more details on demand.}

\begin{figure}
    \centering 
    \includegraphics[width=0.45\linewidth]{boxplot.pdf}
    \caption{Enhanced boxplot of the clusters' characteristics allows a quick
    comparison to the other clusters.\label{fig:boxplot}}
\end{figure}

With the knowledge of what clusters are present, we can explore where they are
through the maps, marked as \emph{where} in Figure~\ref{fig:ui}. While the maps
on the lower part of the interface show the location of the clusters at each
time, the larger map shows a summarised version, based on the trajectories of
each region. \revision{1.9}{In simplified mode the colour corresponds to the
cluster present in that trajectory for more than half the temporal samples, and
is grey otherwise. With simplified colours disabled, the colours are the average
of the colours of the clusters in that trajectory in RGB colour
space~\footnote{\url{https://gka.github.io/chroma.js/\#chroma-average}}.}

\revision{2.5}{Each trajectory represents a different temporal path through the
clusters. Therefore the trajectories encode how the demographic characteristics
changed over time. These changes are summarised by the two sections identified
as \emph{how} in Figure~\ref{fig:ui}. In the left panel, there is a Sankey
diagram where each ribbon represents a different trajectory. The widths are
proportional to the population involved.} In our example in Figure~\ref{fig:ui},
the orange and green clusters contain most of the population and are fairly
stable over time. The pink cluster is small and mostly stable. The purple
cluster is increasing, mostly by incorporating areas that were previously
orange. Since the purple group corresponds to the emergent 'young urban' group,
this corroborates the findings of Delmelle~\citep{Delmelle2016,Delmelle2017},
showing that our network-based method can recover results from the traditional
data processing approach.

\revision{2.5}{In the next panel, illustrated in the top middle of
Figure~\ref{fig:ui}, is a transition matrix between the clusters. It indicates a
rounded percentage of regions whose area changed between each pair of
clusters. In this example, we can see that all clusters are fairly stable, with
more than ninety percent of identity transitions.} This kind of table can be
found in the related literature~\citep{Delmelle2016}, so it is familiar to the
advanced users. It not only informs the proportional changes, but allows the
selection of specific changes for further analysis.

The bottom part of the interface contains the details for the selected
trajectories, or for the whole city if nothing is selected, marked as \emph{who}
in Figure~\ref{fig:ui}. The stacked bar plots summarise the overall demographic
composition of these regions. In this example, the maps show the transition from
orange to green and purple in several regions over time. Each aspect is
represented by a stacked bar plot, where the width of each rectangle corresponds
to \revision{1.10}{the percentage of population in that variable over the
considered period. In this case, a little less than half of the population are
married, and the percentage that are Widowers or Divorced is roughly similar.
About half of the population work in Administrative jobs, a third never
completed high-school, approximately forty percent have gross family income
below 50,000USD per year. About sixty percent identify as white. Placing the
mouse over one of the bars will open a small panel with the temporal evolution
of the population percentage of that specific variable, and clicking on the
chevron on the right side expands the corresponding aspect, showing census tract
level statistics, with details of the temporal evolution of each variable and
also the corresponding IQRs for the whole city. Those values differ from the
total population percentage. While the overall population of this region is
about sixty percent White, the average percentage of White population over the
CTs is a little over forty percent.}

>>>>

\begin{figure}
    \centering 
    \includegraphics[width=0.95\linewidth]{overview.pdf}
    \caption{Overview of the proposed method. 
        \textbf{(a)} A network is generated from the original data,
        encoding the changing geographical information. 
        \textbf{(b)} The network is partitioned into an hierarchy. 
        \textbf{(c)} The characteristics and evolution of the clusters are then
        visually represented.
        \label{fig:overview}}
\end{figure}

@@

Each area of each census year is represented as a node, and edges are placed
between nodes if the corresponding regions share geographic borders in the same
year. Further, edges are placed between nodes if the corresponding regions
belong to sequential years and their geometries intersect.
To avoid spurious connections caused by small fluctuations, one of the
geometries is slightly shrunk before the intersection, using a buffer of -1e-6
degrees. While the result would not change considerably by adding more
connections between nearby regions, these extra connections would greatly
increase the computational cost. More importantly, while weights will be
associated with these edges before they are processed, they are not derived from
the geometry, but from the data. The actual intersection area is not considered
in this representation.  This approach leads to a single network representing
the whole spatio-temporal space of the data.

\subsection{Geographic content clustering}
Having tied the regions together into a network, we can now partition it to
identify similar sets of regions. In essence, we are performing
temporal regionalisation by applying a clustering method over this
spatio-temporal network.

@@

Once the distances are associated to the edges, we use watershed
cuts~\citep{Cousty2009} to create an initial clustering, which is then refined
into a hierarchy using the Sorted Maximal Matching (SMM)~\citep{markus2017}. The
initial watershed step is performed to create an initial clustering and reduce
the running time of the SMM. We introduced one new parameter to this method: a
maximum distance threshold for the merges, to avoid the early merging of
outliers. We refer the reader to the original paper~\citep{markus2017} for more
details, including a complete performance evaluation using several metrics. We
chose this algorithm because it is fast, simple, and easily customisable, but
our methodology should work with any hierarchical network clustering algorithm.
However, we caution against the use of single-linkage methods due to their
tendency to form chain-like clusters with elements that are similar pairwise,
but significantly different on the ends. That is also the reason we did not use
a Watershed hierarchy directly~\citep{najman2011equivalence}.

Each resulting cluster is contiguous in the network. This means that two
similar, but non-contiguous, sets of areas will be classified into two different
clusters, which can be counter-intuitive. To overcome this issue, we
\emph{augment} the network with two new edges per node from a nearest neighbours
graph~\citep{scikit-learn} using only the distances between the histograms.
These edges connect nodes with similar content, if they are not already
connected, providing a path for the algorithm to group similar nodes. The
regions connected by those edges will be merged on the first stages of the
clustering, since the nodes they connect are, by definition, as similar as
possible\footnote{This assertion relies on the euclidean distance and its
relationship with the space where k-nn operates.}, leaving the remaining steps
of the hierarchy to be determined only by the geographical edges. We
experimented with different numbers of augmentation edges, but the results were
not consistent, since the distribution of the edges is data dependent. Adding
two edges per node was the smallest number of edges that led to stable and
consistent results in the scenarios available in our prototype. Since the
problem of balancing the data space with the geographical space is relevant for
geographical data analysis, this idea potentially warrants further exploration,
beyond the scope of this work.

@@

We also use these statistical measurements to discover what characterises each
cluster, that is, what makes it different from the others.  We define the
\emph{relevance} of a variable of an aspect based on the distance between the
interquartile ranges (IQR) of the clusters in the same hierarchical level. If
the IQRs overlap for all clusters, that variable is not relevant to the
characterisation of the cluster, but if the IQRs are distant, it means that this
specific range of values is something that only occurs in this cluster.
Examining IQRs therefore provides users a straightforward visual method for
determining what variables most clearly define a given cluster.

@@

\subsection{Clusters and trajectories}
While the partition of the data into different clusters helps the user to
understand what groups exist and where they are, we are also interested in the
evolution of these groups. To examine this process of evolution directly, we
introduce the concepts of \emph{temporal paths} and \emph{trajectories}.

From network theory, a \emph{path} is a sequence of nodes of the network that
are connected pairwise by edges. As each node has associated temporal
information, we call \emph{temporal} any path such that the temporal information
increases along it. For instance, in Figure~\ref{fig:intuition}, the paths ACF,
ADF, ADG, and BEH are temporal paths, but not HEB, CDE, nor FDA. With harmonised
data, as illustrated by the path BEH, the time-series of each region would form
a temporal path, each node would be connected only to its older and newer
versions, and would belong to only one temporal path. Since our data is not
harmonised, more connections are allowed and each node can belong to an
arbitrary number of paths.

@@

Since each node in the sequence that forms the temporal paths has an associated
cluster, we can classify the paths based on the sequence of clusters. We call
each unique sequence of clusters present in this result a \emph{trajectory}.
Regions on the same trajectory had the same sequence of clusters, therefore had
similar temporal evolution.

@@

\begin{figure*}
    \centering 
    \includegraphics[width=0.95\linewidth]{ui.pdf}
    \caption{Initial interface of our method showing the demographic evolution of Chicago and identifying the objectives of each plot.
        \textbf{(what)}: Cluster overview illustrating the most relevant aspect for each cluster. 
        \textbf{(where)}: Maps illustrating the geographical location, both as an overall summary and for each year.
        \textbf{(how)}: Tabular and visual summary of how the regions classification changed over time.
        \textbf{(who)}: Demographic details for the currently selected data.
        \label{fig:ui}}
\end{figure*}

@@

The cluster overview panel, marked as \emph{what} in Figure~\ref{fig:ui},
displays the most relevant aspects for each cluster, based on the distance
between the IQRs, as detailed in Section~\ref{sec:relevance}. This
feature allows the user to quickly understand what exists in this region and
which aspect makes each cluster unique, without going into excessive detail. In
Chicago's example, we expect race to be relevant in general, and indeed, it is
recognised as the most relevant aspect for three of the four most distinct
clusters, with the fourth being education level. As we increase the number of
clusters, they get more nuanced, requiring more than one aspect for their proper
characterisation, as each cluster is divided into its composing sub-clusters.
The \emph{View all} button opens a new panel where all aspects are included,
while the chevron at the side lets the user expand each cluster separately,
allowing the exploration of more details on demand.

\begin{figure}
    \centering 
    \includegraphics[width=0.425\linewidth]{boxplot.pdf}
    \caption{Enhanced boxplot of the clusters' characteristics allows a quick
    comparison to the other clusters.\label{fig:boxplot}}
\end{figure}

With the knowledge of what clusters are present, we can explore where they are
through the maps, marked as \emph{where} in Figure~\ref{fig:ui}. While the maps
on the lower part of the interface show the location of the clusters at each
time, the larger map shows a summarised version, based on the trajectories of
each region. In simplified mode the colour corresponds to the
cluster present in that trajectory for more than half the temporal samples, and
is grey otherwise. With simplified colours disabled, the colours are the average
of the colours of the clusters in that trajectory in RGB colour
space~\footnote{\url{https://gka.github.io/chroma.js/\#chroma-average}}.

Each trajectory represents a different temporal path through the clusters.
Therefore the trajectories encode how the demographic characteristics changed
over time. These changes are summarised by the two sections identified as
\emph{how} in Figure~\ref{fig:ui}. In the left panel, there is a Sankey diagram
where each ribbon represents a different trajectory. Since the associated
regions have different population numbers, the width of each ribbon is
proportional to the smaller of the populations involved, pairwise. This
approximation works well in general, but leads to small inconsistencies in some
specific cases. For instance, in the results for Miami, which is available in
the supplementary material, the combined width of the ribbons decreases
slightly, despite an increase in the population numbers.

In our example in Figure~\ref{fig:ui}, the orange and green clusters contain
most of the population and are fairly stable over time. The pink cluster is
small and mostly stable. The purple cluster is increasing, mostly by
incorporating areas that were previously orange. Since the purple group
corresponds to the emergent 'young urban' group, this corroborates the findings
of Delmelle~\citep{Delmelle2016,Delmelle2017}, showing that our network-based
method can recover results from the traditional data processing approach.

In the next panel, illustrated in the top middle of Figure~\ref{fig:ui}, is a
transition matrix between the clusters. It indicates a rounded percentage of
regions whose area changed between each pair of clusters. In this example, we
can see that all clusters are fairly stable, with more than ninety percent of
identity transitions. This kind of table can be found in the related
literature~\citep{Delmelle2016}, so it is familiar to the advanced users. It not
only informs the proportional changes, but allows the selection of specific
changes for further analysis.

The bottom part of the interface contains the details for the selected
trajectories, or for the whole city if nothing is selected, marked as \emph{who}
in Figure~\ref{fig:ui}. The stacked bar plots summarise the overall demographic
composition of these regions. In this example, the maps show the transition from
orange to green and purple in several regions over time. Each aspect is
represented by a stacked bar plot, where the width of each rectangle corresponds
to the percentage of population in that variable over the considered period. In
this case, a little less than half of the population are married, and the
percentage that are Widowers or Divorced is roughly similar. About half of the
population work in Administrative jobs, a third never completed high-school,
approximately forty percent have gross family income below 50,000USD per year.
About sixty percent identify as white. Placing the mouse over one of the bars
will open a small panel with the temporal evolution of the population percentage
of that specific variable, and clicking on the chevron on the right side expands
the corresponding aspect, showing census tract level statistics, with details of
the temporal evolution of each variable and also the corresponding IQRs for the
whole city. Those values differ from the total population percentage. While the
overall population of this region is about sixty percent White, the average
percentage of White population over the CTs is a little over forty percent.

### 2019-10-10 paper/related.tex
last paper review piccard 1.0 GEAN

Since our problem encompasses several fields, we divide this section into
specific sub problems: \emph{longitudinal demographic studies}, describing the
traditional tabular approach to longitudinal studies; \emph{data
representation}, elaborating how evolving geographic data can be represented for
processing; \emph{data clustering and regionalisation}, briefly reviewing
existing data clustering methods,\revision{2.1}{ geographic constraints and
regionalisation}; and \emph{cluster characterisation}, articulating how clusters
can be visually summarised.

\subsection{Longitudinal demographic studies}
Census data is used not only to discover demographic
patterns~\citep{Firebaugh2016}, but to correlate demographic characteristics to
other measurements~\citep{diez1997neighborhood}. However, longitudinal studies
are rare, because they are difficult : \emph{"(...) One of the most challenging
and fascinating areas in spatial statistics is the synthesis of spatial data
collected at different spatial scales(...)"}~\citep{gotway2002combining}. While
census tract level data is readily available for the US since at least
1910~\citep{nhgis}, most studies consider the period between 1970 and 2010,
\revision{1.3}{using pre-harmonised data from the Longitudinal Tract Data
Base~\citep{Logan2014} or the Neighborhood Change Database~\citep{ncdb}, albeit
their inherent errors~\citep{Logan2016,Hallisey2017}.} Similar harmonisation
efforts appear in other countries~\citep{Liu2015,Lee2015,Allen2018}. These
datasets have been highly significant for the field. Yet they also limit the
universe of data that can be used to study neighbourhood change, since any new
datasets would need to be similarly processed in order to be rendered compatible
with these sources.

Another option considers the use of grid
data~\citep{Dmowska2017,Dmowska2018,stepinski2019imperfect}. Beyond the
potentially increased spatial\revision{1.4}{ precision}, this approach does not
require complex harmonisation when new data is considered, if the grids are
compatible. However, demographic data is usually not available in this format,
especially from older sources. Additionally, the conversion from tabulation
areas can introduce significant errors.

@@

Data clustering is one of the elementary processes for data analysis,
simplifying the data into a smaller number of homogeneous sets that can be
interpreted in the same way. There is no shortage of contributions for this
problem~\citep{Fahad2014}\revision{}{, since variations of it appear in almost all
scientific fields.}

\revision{2.1}{In geography, this problem is known as
\emph{regionalisation}~\citep{montello2003regions}, a rather old problem that
has been thoroughly explored, leveraging different mathematical tools, including
discrete topology~\citep{brantingham1978topological} and discrete
geometry~\citep{assunccao2006efficient}. Indeed, network-based methods are among
the current state-of-the-art~\citep{Guo2008Redcap,maxp}. However,
\emph{temporal} regionalisation is significantly less explored, especially in a
demographic context, arguably due to the difficulties in dealing with
unharmonised longitudinal data. Recent neighbourhood related applications rely
on k-means~\citep{jain2010data,Delmelle2016}, the Louvain method for community
detection~\citep{blondel2008Louvain,Thomas2012}, or, to a lesser extent, Self
Organising Maps~\citep{Delmelle2017,Ling2016,Dani2013_som_maxp}.}

\revision{1.5/2.2}{Since we adopted a network-based data representation and our objectives include
an interactive interface, we opted for an heuristic variation of the maximum
weighted matching algorithm called \emph{sorted maximal
matching}~\citep{markus2017}, because of its simplicity, customisability, and
fast computation times. This algorithm operates on weighted networks, where a
distance metric between the nodes is associated with the edges. We adopted a
distance based on the data associated with each node. The algorithm merges
clusters based on these distances, creating a hierarchy instead of a fixed
number of clusters. This hierarchical result is rendered by the interface,
allowing the user to change the number of clusters without reprocessing.
Changing the clustering algorithm would lead to different results, but any
hierarchical network clustering method, and distance metric, can be used in our
framework. }

>>>>

Since our problem encompasses several fields, we divide this section into
specific sub problems: \emph{longitudinal demographic studies}, describing the
traditional tabular approach to longitudinal studies; \emph{data
representation}, elaborating how evolving geographic data can be represented for
processing; \emph{data clustering and regionalisation}, briefly reviewing
existing data clustering methods, geographic constraints and
regionalisation; and \emph{cluster characterisation}, articulating how clusters
can be visually summarised.

\subsection{Longitudinal demographic studies}
Census data is used not only to discover demographic
patterns~\citep{Firebaugh2016}, but to correlate demographic characteristics to
other measurements~\citep{diez1997neighborhood}. However, longitudinal studies
are rare, because they are difficult : \emph{"(...) One of the most challenging
and fascinating areas in spatial statistics is the synthesis of spatial data
collected at different spatial scales(...)"}~\citep{gotway2002combining}. While
census tract level data is readily available for the US since at least
1910~\citep{nhgis}, most studies consider the period between 1970 and 2010,
using pre-harmonised data from the Longitudinal Tract Data
Base~\citep{Logan2014} or the Neighborhood Change Database~\citep{ncdb}, despite
their inherent errors~\citep{Logan2016,Hallisey2017}. Similar harmonisation
efforts appear in other countries~\citep{Liu2015,Lee2015,Allen2018}. These
datasets have been highly significant for the field. Yet they also limit the
universe of data that can be used to study neighbourhood change, since any new
datasets would need to be similarly processed in order to be rendered compatible
with these sources.

Another option considers the use of grid
data~\citep{Dmowska2017,Dmowska2018,stepinski2019imperfect}. Beyond the
potentially increased spatial precision, this approach does not require complex
harmonisation when new data is considered, if the grids are compatible. However,
demographic data is usually not available in this format, especially from older
sources. Additionally, the conversion from tabulation areas can introduce
significant errors.

@@

Data clustering is one of the elementary processes for data analysis,
simplifying the data into a smaller number of homogeneous sets that can be
interpreted in the same way. There is no shortage of contributions for this
problem~\citep{Fahad2014}, since variations of it appear in almost all
scientific fields.

In geography, this problem is known as
\emph{regionalisation}~\citep{montello2003regions}, a rather old problem that
has been thoroughly explored, leveraging different mathematical tools, including
discrete topology~\citep{brantingham1978topological} and discrete
geometry~\citep{assunccao2006efficient}. Indeed, network-based methods are among
the current state-of-the-art~\citep{Guo2008Redcap,maxp}. However,
\emph{temporal} regionalisation is significantly less explored, especially in a
demographic context, arguably due to the difficulties in dealing with
unharmonised longitudinal data. Recent neighbourhood related applications rely
on k-means~\citep{jain2010data,Delmelle2016}, the Louvain method for community
detection~\citep{blondel2008Louvain,Thomas2012}, or, to a lesser extent, Self
Organising Maps~\citep{Delmelle2017,Ling2016,Dani2013_som_maxp}.

Since we adopted a network-based data representation and our objectives include
an interactive interface, we opted for an heuristic variation of the maximum
weighted matching algorithm called \emph{sorted maximal
matching}~\citep{markus2017}, because of its simplicity, customisability, and
fast computation times. This algorithm operates on weighted networks, where a
distance metric between the nodes is associated with the edges. We adopted a
distance based on the data associated with each node. The algorithm merges
clusters based on these distances, creating a hierarchy instead of a fixed
number of clusters. This hierarchical result is rendered by the interface,
allowing the user to change the number of clusters without reprocessing.
Changing the clustering algorithm would lead to different results, but any
hierarchical network clustering method, and distance metric, can be used in our
framework.
